Do moral communities have a spatial dimension? A spatial exploratory analysis of places of worship and violent crime in the city of Recife, Brazil

### 2019-10-15 Moral_Communities_and_Crime_v1.Rmd
First complete draft of paper and title page

Words go here.

>>>>

Theories of moral communities and social disorganization suggest that the presence of institutions that can provide a measure of social control can help to improve crime prevention, and this has been borne to some extent by previous empirical research. However, as noted by Warner and Konkel [-@Warner2019neighborhood], it is important to distinguish between different religious traditions.

In this respect, it has been argued that Catholics and some Protestant groups are more adept at generating bridging social capital with stron intergroup relations. This can result in greater collective efficacy - which in turn can decrease the likelihood of crime [see @Hoffmann2015religion, p. 2]. Along these lines, Beyerlein and Hipp [-@Beyerlein2005social], examined data for 3,157 counties in the United States in 2000, and found that counties with greater percentages of people affiliated with mainline Protestant and Catholic traditions tended to have lower crime rates. Similarly, Warner and Konkel [-@Warner2019neighborhood] found that mainline Protestant and briding churches (but not Catholic churches) correlated positively with informal social control at the neighborhood level in two cities in Kentucky. Nie and Yang [-@Nie2019smoking], in a study of smoking, also found that share of Catholic population in counties did not reduce the rate of smoking, thus bringing into question the efficacy of this type of moral community in generating social control - albeit at a fairly aggregated scale (counties). Our findings with respect to Catholic places of worship echo these results at a much higher level of disaggregation, as we fail to find evidence that Catholic churches project a moral community in space more than any of various common types of business establishments.

Conservative Protestant and Evangelic traditions, unlike mainline Protestants and Catholics, appear to focus less on bridging and more on bonding social capital with strong within-group relations. Associated with this, Conservative Protestant doctrine emphasizes individual responsibility, and therefore social ills tend to be seen as primarily personal ills in need of religious redemption instead of secular interventions [@Nie2019smoking, pp. 2-3]. Under this light, a moral community with such views of social problems may not be effective to deter crime [@Ellison2003enduring]; instead, conservative Protestant doctrine may be more tolerant of violent behavior when associated with defense of honor, family, property, or women, or unfortunate events may be seen as representing legitimate celestial retribution for moral torpitude. Empirical research by Beyerlein and Hipp [-@Beyerlein2005social] found that counties with greater percentages of people affiliated with Evangelical traditions tended to have higher crime rates. More recent research by Desmond et al. [-@Desmond2010congregations] in more than 400 block groups in Indianapolis found that neighborhoods with a greater presence of Evangelical congregations had higher rates of both violent and property crimes. In the specific context of Brazil, it is important to consider the role that religion has had in interpreting and conferring meaning to urban violence, or what Brazilian anthropologist Patricia Birman has termed "The Violence of the Just" [see @Birman2012violencia; @Birman2019narrativas].

Our analysis suggests that the role of Evangelical churches may be more complex than thought: on the one hand, our results indicate that Lethal and Intentional Violent Crime tends to be more intense in the proximity of Evangelical places of worship, up to a distance of approximately 100 m, but then declines even below the global intensity and the background intenstiy at distances between 300 m and 100 m. It is at the moment unclear what could cause this, and candidate explanations could include the way police respond to crime incidents in different neighborhoods [e.g., Warner, 1997], the likelihood of residents reporting crime to the police (e.g., Goudriaan, Witterbrood, & Niewbeerta, 2006), and the level of opportunity for crimes (e.g., Rountree, Land, & Miethe, 1994). Could Evangelicals churches provide greater levels of opportunity for crime, if adherents to this tradition are seen as meek? And is it possible that at a different scale, unlike reports by other researchers, Evangelical churches are more effective at keeping social control? While these speculations cannot be answered in the context of the present study, we suggest that these might be fruitful areas for future research.

Finally, we would like to remark on the results regarding Spiritist churches, which in some way are the mirror image of those for Evangelical churches. An extensive scan of the literature finds much less information about this religious tradition, in particular in the context of Brazil where it adapted in ways that made it very different from its original French predecessor [see @daGraca2011espiritismo]. Could it be that Spiritism is seen by potential criminals as spiritually or materially risky, if supernatural retribution is feared? Again, we cannot do more than speculate about this at the moment, but this might be another worthwhle avenue for future investigation.

### 2019-10-15 Moral_Communities_and_Crime_v1.tex
First complete draft of paper and title page

Words go here.

@@

@@

@@

@@

>>>>

Theories of moral communities and social disorganization suggest that
the presence of institutions that can provide a measure of social
control can help to improve crime prevention, and this has been borne to
some extent by previous empirical research. However, as noted by Warner
and Konkel (2019), it is important to distinguish between different
religious traditions.

In this respect, it has been argued that Catholics and some Protestant
groups are more adept at generating bridging social capital with stron
intergroup relations. This can result in greater collective efficacy -
which in turn can decrease the likelihood of crime (see Hoffmann 2015,
2). Along these lines, Beyerlein and Hipp (2005), examined data for
3,157 counties in the United States in 2000, and found that counties
with greater percentages of people affiliated with mainline Protestant
and Catholic traditions tended to have lower crime rates. Similarly,
Warner and Konkel (2019) found that mainline Protestant and briding
churches (but not Catholic churches) correlated positively with informal
social control at the neighborhood level in two cities in Kentucky. Nie
and Yang (2019), in a study of smoking, also found that share of
Catholic population in counties did not reduce the rate of smoking, thus
bringing into question the efficacy of this type of moral community in
generating social control - albeit at a fairly aggregated scale
(counties). Our findings with respect to Catholic places of worship echo
these results at a much higher level of disaggregation, as we fail to
find evidence that Catholic churches project a moral community in space
more than any of various common types of business establishments.

Conservative Protestant and Evangelic traditions, unlike mainline
Protestants and Catholics, appear to focus less on bridging and more on
bonding social capital with strong within-group relations. Associated
with this, Conservative Protestant doctrine emphasizes individual
responsibility, and therefore social ills tend to be seen as primarily
personal ills in need of religious redemption instead of secular
interventions (Nie and Yang 2019, 2--3). Under this light, a moral
community with such views of social problems may not be effective to
deter crime (Ellison, Burr, and McCall 2003); instead, conservative
Protestant doctrine may be more tolerant of violent behavior when
associated with defense of honor, family, property, or women, or
unfortunate events may be seen as representing legitimate celestial
retribution for moral torpitude. Empirical research by Beyerlein and
Hipp (2005) found that counties with greater percentages of people
affiliated with Evangelical traditions tended to have higher crime
rates. More recent research by Desmond et al.~(2010) in more than 400
block groups in Indianapolis found that neighborhoods with a greater
presence of Evangelical congregations had higher rates of both violent
and property crimes. In the specific context of Brazil, it is important
to consider the role that religion has had in interpreting and
conferring meaning to urban violence, or what Brazilian anthropologist
Patricia Birman has termed ``The Violence of the Just'' (see Birman and
Machado 2012; Birman 2019).

Our analysis suggests that the role of Evangelical churches may be more
complex than thought: on the one hand, our results indicate that Lethal
and Intentional Violent Crime tends to be more intense in the proximity
of Evangelical places of worship, up to a distance of approximately 100
m, but then declines even below the global intensity and the background
intenstiy at distances between 300 m and 100 m. It is at the moment
unclear what could cause this, and candidate explanations could include
the way police respond to crime incidents in different neighborhoods
{[}e.g., Warner, 1997{]}, the likelihood of residents reporting crime to
the police (e.g., Goudriaan, Witterbrood, \& Niewbeerta, 2006), and the
level of opportunity for crimes (e.g., Rountree, Land, \& Miethe, 1994).
Could Evangelicals churches provide greater levels of opportunity for
crime, if adherents to this tradition are seen as meek? And is it
possible that at a different scale, unlike reports by other researchers,
Evangelical churches are more effective at keeping social control? While
these speculations cannot be answered in the context of the present
study, we suggest that these might be fruitful areas for future
research.

Finally, we would like to remark on the results regarding Spiritist
churches, which in some way are the mirror image of those for
Evangelical churches. An extensive scan of the literature finds much
less information about this religious tradition, in particular in the
context of Brazil where it adapted in ways that made it very different
from its original French predecessor (see Graca Arribas 2011). Could it
be that Spiritism is seen by potential criminals as spiritually or
materially risky, if supernatural retribution is feared? Again, we
cannot do more than speculate about this at the moment, but this might
be another worthwhle avenue for future investigation.

@@

\leavevmode\hypertarget{ref-Beyerlein2005social}{}%
Beyerlein, Kraig, and John R Hipp. 2005. ``Social Capital, Too Much of a
Good Thing? American Religious Traditions and Community Crime.'' Journal
Article. \emph{Social Forces} 84 (2): 995--1013.

\leavevmode\hypertarget{ref-Birman2012violencia}{}%
Birman, Patricia, and Carly Machado. 2012. ``A Violencia Dos Justos:
Evangelicos, Midia E Periferias Da Metropole.'' \emph{Revista Brasileira
de Ciências Sociais} 27 (80): 55--69.

\leavevmode\hypertarget{ref-Birman2019narrativas}{}%
Birman, Patrocia. 2019. ``NARRATIVAS Seculares E Religiosas Sobre a
Violencia: AS Fronteiras Do Humano No Governo Dos Pobres.''
\emph{Sociologia \& Antropologia} 9 (1).

@@

\leavevmode\hypertarget{ref-Desmond2010congregations}{}%
Desmond, S. A., G. Kikuchi, and K. H. Morgan. 2010. ``Congregations and
Crime: Is the Spatial Distribution of Congregations Associated with
Neighborhood Crime Rates?'' Journal Article. \emph{Journal for the
Scientific Study of Religion} 49 (1): 37--55.
\url{https://doi.org/10.1111/j.1468-5906.2009.01491.x}.

@@

\leavevmode\hypertarget{ref-Ellison2003enduring}{}%
Ellison, Christopher G, Jeffrey A Burr, and Patricia L McCall. 2003.
``The Enduring Puzzle of Southern Homicide: Is Regional Religious
Culture the Missing Piece?'' \emph{Homicide Studies} 7 (4): 326--52.

@@

\leavevmode\hypertarget{ref-daGraca2011espiritismo}{}%
Graca Arribas, Celia da. 2011. ``Espiritismo: Entre Crime E Religiao.''
\emph{Mneme-Revista de Humanidades} 12 (29).
